---
title: "A2S Example: Service accessibility for user specified LSOA in Wales, UK"
output:
  pdf_document: default
  html_notebook: default
---

# Specific LSOA area you want to run for

```{r message=FALSE, warning=FALSE}
lsoa_name = 'Cardiff'
service_type = 'atm' #atm, pub, school, pharmacy etc (https://wiki.openstreetmap.org/wiki/Key:amenity)
```

# Setup connection to Open Trip Planner

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(sf)
library(propeR)

#setup connection to the OTP server
otpcon <- otpConnect()

#output dir
out_dir = paste(getwd(),"output", sep="/")
dir.create(out_dir, showWarnings = FALSE)
```

# Read in OpenStreetMap data using Overpass

```{r message=FALSE, warning=FALSE}

library(osmdata)

service_overpass <- opq(paste0(lsoa_name," Wales")) %>% # Area bounding box
  add_osm_feature(key = 'amenity', value = service_type, value_exact = FALSE) %>%
  osmdata_sf()
services <- service_overpass$osm_points
coords = sf::st_coordinates(services)
#propeR requires lon, lat columns
services['lon'] = coords[,'X']
services['lat'] = coords[,'Y']
services['lat_lon'] = paste(services$lat, services$lon, sep=",")

services <- services %>% 
  dplyr::select(name=osm_id, lat, lon, lat_lon)

row.names(services) <- services$name

```

# Read in LSOA data from ONS geography portal

```{r message=FALSE, warning=FALSE}
# Want lsoas for visualisation, but population weighted centroids for analysis
#Read in both using the geoportal API

lsoas <- sf::st_read(paste0('https://services1.arcgis.com/ESMARspQHYMw9BZ9/arcgis/rest/services/LSOA_DEC_2011_EW_BFE/FeatureServer/0/query?where=LSOA11NM%20like%20%27%25',toupper(lsoa_name),'%25%27&outFields=LSOA11CD,LSOA11NM,FID&outSR=4326&f=json'), 
                     stringsAsFactors=F, 
                     quiet=T)

lsoa_pop_centroids <- sf::st_read(paste0('https://ons-inspire.esriuk.com/arcgis/rest/services/Census_Boundaries/Lower_Super_Output_Areas_December_2011_Centroids/MapServer/0/query?where=UPPER(lsoa11nm)%20like%20%27%25',toupper(lsoa_name),'%25%27&outFields=*&outSR=4326&f=json'),
                                  stringsAsFactors=F, 
                                  quiet=T) 

# Solves discrepancy amongst column names
names(lsoa_pop_centroids)[2] <- "LSOA11CD"

#seperate out the centroid coords 
coords = lsoa_pop_centroids %>% sf::st_coordinates()
lsoa_pop_centroids['lon'] = coords[,'X']
lsoa_pop_centroids['lat'] = coords[,'Y']

lsoa_pop_centroids <- lsoa_pop_centroids %>%
  dplyr::mutate(lat_lon = paste(lat, lon,sep=",")) %>%
  sf::st_drop_geometry() %>%
  dplyr::select(LSOA11CD, lat, lon, lat_lon) 

lsoas <- lsoas %>%
  dplyr::left_join(lsoa_pop_centroids, by="LSOA11CD") %>%
  dplyr::select(LSOA11CD, name = LSOA11NM, lat, lon, lat_lon)
```

# Read in NOMIS data

```{r message=FALSE, warning=FALSE}
#read in 2017 population estimates from Nomis and add them onto the lsoas

population_nomis = paste0("http://www.nomisweb.co.uk/api/v01/dataset/NM_2010_1.data.csv?geography=",paste(lsoas$LSOA11CD, collapse = ","),"&date=latest&gender=0&c_age=200&measures=20100&select=geography_code,obs_value")

population = readr::read_csv(population_nomis, col_names=c('LSOA11CD', 'pop_2017'), col_types = "ci", skip = 1) 

lsoas <- lsoas %>% dplyr::left_join(population, by ='LSOA11CD')

row.names(lsoas) <- lsoas$LSOA11CD

```

# Clean location data

```{r message=FALSE, warning=FALSE}
services <- locationValidatorIsochrone(output.dir = out_dir,
                                       otpcon = otpcon, locationPoints = services)
lsoas <- locationValidatorIsochrone(output.dir = out_dir,
                                       otpcon = otpcon, locationPoints = lsoas)
```


# Calculate the isochrones around the Services

```{r message=FALSE, warning=FALSE}

# run propeR using services as the origins, calculate bands from 5 mins to half an hour in 5 min steps
# assume they are walking.

# it can be that the origins are too far from the road network and so it doesn't find a route, the 
# locationvalidator can be used to resolve a location in that case.

filename <- propeR::isochroneMulti(output.dir = out_dir,
                                   otpcon = otpcon,
                                   originPoints = as.data.frame(services),
                                   destinationPoints = as.data.frame(lsoas),
                                   startDateAndTime  = '2018-08-18 12:00:00',
                                   modes = 'WALK',
                                   isochroneCutOffMax = 30,
                                   isochroneCutOffMin = 10,
                                   isochroneCutOffStep = 5,
                                   mapOutput = F,
                                   originMarker = F,
                                   geojsonOutput = T)
```

# Union the isochrones for individual time bands and difference to get distinct areas

```{r message=FALSE, warning=FALSE}

# currently you need to find the prefix of the output file somehow - I've just taken it from the filesystem
isoc_output <- sf::st_read(paste0(out_dir,"/",filename,"/geojson/", filename,".geoJSON"), quiet=T)

# an isochrone will have been created for each Service and time step, so union together the set of time steps to get
# one, or a multigeom, for each step
isoc_union <- isoc_output %>% 
  sf::st_buffer(0.0) %>% #helps resolve self intersections
  dplyr::group_by(time) %>%
  dplyr::summarise() %>%
  dplyr::arrange(desc(time)) 

# The time bands include all areas up to that time, want bands that just cover the step e.g. 20-30mins
# So progressively difference between pairs of bands
for (i in seq(nrow(isoc_union)-1)){
  sf::st_geometry(isoc_union[i,]) <- sf::st_difference(isoc_union$geometry[i],isoc_union$geometry[i+1])
}

```

# Plot the result using Leaflet

```{r message=FALSE, warning=FALSE}
library(leaflet)
library(RColorBrewer)

#convert to mins
isoc_union$time_mins = (isoc_union$time / 60)

#using colorbrewer sequential palette for time bands
bins <- c(0,sort(isoc_union$time_mins))
pal <- leaflet::colorBin("PuBu", domain = isoc_union$time_mins, bins = bins,reverse = T)

#plot lsoas (for context), time bands, and Service locations
leaflet::leaflet(isoc_union) %>%
  leaflet::addScaleBar() %>%
  leaflet::addProviderTiles(providers$CartoDB.Positron) %>%
  leaflet::addPolygons(data = lsoas, 
                       fill = F, 
                       weight = 1,
                       color = "orange",
                       group="lsoas") %>%
  leaflet::addPolygons(
    fillColor = ~pal(time_mins),
    opacity = 1,
    weight = 1,
    color = 'grey',
    fillOpacity = 0.75,
    group = 'isochrones'
  ) %>% 
  leaflet::addCircleMarkers(lat = services$lat, 
                            lng = services$lon,
                            color = 'red',
                            stroke = T, 
                            fill=F,
                            radius = 2, 
                            weight = 1,
                            fillOpacity = 0.8, 
                            label = service_type
  ) %>%
  leaflet::addLegend(pal = pal, 
                     values = ~time_mins, 
                     opacity = 0.7, 
                     title = "Time to nearest Service (mins)",
                     position = "bottomright"
  )
```

# Calculate accessibility statistics

```{r message=FALSE, warning=FALSE}
library(lwgeom)
library(formattable)

# Estimate the population covered by each time band
# using an apportionment by area for calcuation 
# Need to use an intersection rather than a spatial join to get sub-areas

pop_coverd <- lsoas %>% 
  dplyr::mutate(lsoa_area = sf::st_area(geometry)) %>%
  dplyr::select(LSOA11CD, pop_2017, lsoa_area) %>%
  sf::st_intersection(isoc_union) %>%
  dplyr::mutate(seg_area = sf::st_area(geometry),
                prop_area = seg_area / lsoa_area, 
                prop_pop = as.numeric(prop_area * pop_2017)
  ) %>%
  sf::st_drop_geometry() %>%
  dplyr::group_by(time_mins) %>%
  dplyr::summarise(pop_covered = sum(prop_pop)) %>%
  dplyr::ungroup() 

# Intersection loses areas not covered by bands so need to add population for those back in
pop_remaining = sum(lsoas$pop_2017) - sum(pop_coverd$pop_covered)

#Plot result in a table
pop_coverd %>% 
  dplyr::mutate(time_mins = as.factor(time_mins)) %>%
  dplyr::bind_rows(tibble(time_mins = "> 30", pop_covered = pop_remaining)) %>%
  dplyr::mutate(pc_pop = formattable::percent(pop_covered / sum(pop_covered), digits=1),
                pop_covered = formattable::digits(pop_covered,digits = 0)) %>%
  formattable::formattable(col.names = c("Time (mins)", "Population", "% Population"))
```

